Appendix 3: Estimating species interactions

### 2015-11-02 Harris_2015/manuscript-materials/draft.Rmd
etc

**Title:** Estimating species interactions from co-occurrence data with Markov
networks

@@

**Abstract:** Estimating species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can also be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known pairwise interaction strengths, I evaluated Markov
networks and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions between
species pairs even when indirect interactions or abiotic factors largely
overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = -2$, then each species' odds of occurrence are
$exp{2}$ times lower when the other one is present (as compared with otherwise
comparable sites). Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else equal---to
produce assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an ecologist
might expect). Of course, if all else is *not* equal (e.g. Figure 1, where the
presence of one competitor is associated with release from another competitor),
then species' overall co-occurrence patterns can be more complicated, and may
require summing over the different possible assemblages (Figure 2B) to determine
the overall outcomes.

@@

\noindent \textbf{Simulated landscapes.}
I simulated several sets of landscapes using known parameters as a test bed for
different statistical methods. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling from a probability distribution defined by
the figure's interaction coefficients (Appendix 1). Each of the methods
described below was then evaluated on its ability to correctly infer that the
two shrub species competed with one another, despite their frequent co-occurrence.

@@

\noindent
\textbf{Twenty species.}
Despite some variability across contexts (Figure 3), the four methods that
controlled for indirect effects (the Markov network, the generalized linear
models, and the two partial correlation-based methods) outperformed those
that did not.  Of these, the Markov network consistently performed the best.
As anticipated by @lee_learning_2012, the Markov network produced very
similar estimates to the generalized linear models (Figure 4A), but with
higher accuracy---especially with smaller data sets.  As reviewed in
@gotelli_empirical_2009, most analyses in this field in ecology involve
fewer than 50 sites, so this is the most important regime for model
performance. As shown in Appendix [[X]], the standard errors associated
with the estimates in Figure 3 are small (less than 0.01), so these
differences should not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). The estimates from these three approaches
were tightly correlated, after controlling for the size of the landscape,
suggesting that the null model only contains a noisy version of the
same information that could be obtained more easily and interpretably with
simple correlation coefficients (Figure 4B).

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's Type I error rate (defined here as its
probability of rejecting the null hypothesis for a pair of species where
$|\beta_{ij}|<0.1$ depended on the simulation type: 0.03 for simulations that
matched the model's assumptions, versus 0.15 for simulations that included
environmental heterogeneity [[Appendix]].  For the null model, these rates were
0.31 and 0.50, respectively. Figure 4C shows, across a range of true interaction
strengths, the probability that the null model or the Markov network will
predict the wrong sign of the interaction with 95% confidence.  The null model
makes such errors nearly 8 times as often as the Markov network, even though it
only reject the null hypothesis twice as often overall [[Appendix]].  The Markov
network's errors were also more concentrated around 0, so it never misclassified
strong interactions like the null model did.

@@

\noindent
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
environmental heterogeneity and indirect interactions cause ecologists' typical
null modeling approaches to reliably fail. Partial covariances and generalized
linear models can both provide computationally efficient approximations, but
with somewhat lower accuracy, especially for typically-sized data sets with small
numbers of sites [@gotelli_empirical_2009]. The difference in accuracy may be
larger for real data sets than for the simulated landscapes in Figure 3,
however; linear approximations to the Markov network make larger errors when
the interaction matrix is structured [e.g. due to guilds or trophic levels;
@loh_structure_2013]. Similarly, the separate generalized linear models
for each species can severely overfit in some cases [@lee_learning_2012]. The
full Markov network should thus be preferred to the approximations when
it is computationally feasible.

Compositional data only contains enough degrees of freedom to estimate one
interaction per species pair, so none of these methods can identify the exact
nature of the pairwise interactions [e.g. which species in a
positively-associated pair is facilitating the other; cf
@schmidt_modeling_2012]. To estimate asymmetric interactions, such as
commensalism or predation, ecologists could use time series, behavioral
observations, manipulative experiments, or natural history. These other sources
of information could also be used to augment the likelihood function with an
informative prior distribution, reducing ecologists' error and uncertainty.

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they are less vulnerable than some
of the most commonly-used methods to mistakenly identifying positive species
interactions between competing species, and can make precise statements about
the conditions where indirect interactions will overwhelm direct ones. They also
provide a simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flashpoint for the null model
debates in the 1980’s [@roughgarden_competition_1983; @strong_ecological_1984].
Equation 1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{e^\alpha}{1 + e^{\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated by
subtracting this value from the observed prevalence (cf Figure 2D). This kind of
insight would have been difficult to obtain without a generative model that
predicts the consequences of species interactions (as opposed to null
simulations that omit interactions entirely).

Markov networks---particularly the Ising model for binary networks---have been
studied for nearly a century [@cipra_introduction_1987], and the models are
well-understood. Using the same framework for species interactions would thus
allow ecologists to tap into a vast set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: for example, the states of the
interaction network can be modeled as a function of the local abiotic
environment [@lee_learning_2012], which would help incorporate networks of
biotic interactions into species distribution models
[@pollock_understanding_2014] and lead to a better understanding of the
interplay between biotic and abiotic effects on community structure.
Alternatively, models could allow one species to alter the relationship between
two other species [@tjelmeland_markov_1998; cf @bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
specific ecological processes like competition by rejecting a general-purpose
null [@gotelli_empirical_2009], especially when the test statistic is
effectively just a correlation coefficient Figure 4A. When a non-null
backdrop is not controlled for, Type I error rates can skyrocket, the apparent
sign of the interaction can change, and null models can lead to reliably
incorrect inferences (Figure 1, Figure 4).

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3A, scientists in this field could more than double
their explanatory power by switching from null models to Markov networks (or
increase it nearly as much with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for indirect
effects could clearly improve our understanding of species' direct effects on
one another and on community structure.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, but more weakly (circled coefficient). **B.** In spite of the
competitive interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors positively
correlated (circled). **C.** Controlling for the tree species' presence with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative shrub-shrub
interaction (circled).

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta_12$ coefficient is consistent with competition between the two species.
**B.** The coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and includes $\beta_12$ only when both are
present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
frequencies of all possible co-occurrence patterns between the two species of
interest, as calculated in the previous panel. **D.** If $\beta_{12}$ equaled
zero (e.g. if the species no longer competed for the same resources), then the
each species' overall occurrence rates would increase.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations.

\noindent \textbf{Figure 4: A.} The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates. **B.** The null
model's estimates typically matched the correlation coefficient, after
controlling for landscape size. **C.** For any given interaction strength, the
null model was much more likely to misclassify its sign with 95% confidence than
the Markov network was.

>>>>

**Title:** Inferring species interactions from co-occurrence data with Markov
networks

@@

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can also be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known pairwise interaction strengths, I evaluated Markov
networks and six existing approaches. The Markov networks consistently
outperformed other methods, correctly isolating direct interactions between
species pairs even when indirect interactions or abiotic factors largely
overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else equal---to
produce assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an ecologist
might expect). Of course, if all else is *not* equal (e.g. Figure 1, where the
presence of one competitor is associated with release from another competitor),
then predicting species' overall co-occurrence rates can be more complicated,
and may require summing over the different possible assemblages (Figure 2B).

@@

\noindent \textbf{Simulated landscapes.}
I simulated several sets of landscapes using known parameters to evaluate
different statistical methods' performance. The first set of landscapes included
the three competing species shown in Figure 1. For each of 1000 replicates, I
generated a landscape with 100 sites by sampling from a probability distribution
defined by the figure's interaction coefficients (Appendix 1). Each of the
methods described below was then evaluated on its ability to correctly infer
that the two shrub species competed with one another, despite their frequent
co-occurrence.

@@

\noindent
\textbf{Twenty species.}
Despite some variability across contexts (Figure 3), the four methods that
controlled for indirect effects (the Markov network, the generalized linear
models, and the two partial correlation-based methods) outperformed those
that did not.  Of these, the Markov network consistently performed the best.
As anticipated by @lee_learning_2012, generalized linear models closely
approximated the Markov network estimates, especially when the data sets were
very large. As reviewed in @gotelli_empirical_2009, however, most analyses in
this field involve fewer than 50 sites, so model performance on the smaller
landscapes where the gap was larger is most important. As shown in Appendix
[[X]], the standard errors associated with the estimates in Figure 3 are small
(less than 0.01), so the differences among methods should not be attributed to
sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). The estimates from these approaches
were tightly correlated (after controlling for the size of the landscape)
suggesting that the null model only contains a noisy version of the
same information that could be obtained more easily and interpretably with
simple correlation coefficients (Figure 4B).

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here as
its probability of rejecting the null hypothesis for a pair of species where
$|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for simulations that
matched the model's assumptions, versus 0.15 for simulations that included
environmental heterogeneity [[Appendix]].  For the null model, these rates were
0.31 and 0.50, respectively---far higher than the nominal 0.05 rate. Figure 4C
shows, across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the interaction
with 95% confidence.  The null model makes such errors nearly 8 times as often
as the Markov network, even though it only reject the null hypothesis twice as
often overall [[Appendix]].  The Markov network's errors were also more
concentrated around 0, so it never misclassified strong interactions like the
null model did.

@@

\noindent
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
environmental heterogeneity and indirect interactions cause ecologists' typical
null modeling approaches to reliably fail. Partial covariances and generalized
linear models can both provide computationally efficient approximations, but
with somewhat lower accuracy, especially for typically-sized data sets with small
numbers of sites [@gotelli_empirical_2009]. The difference in accuracy may be
larger for real data sets than for the simulated landscapes in Figure 3,
however; linear approximations to the Markov network make larger errors when
the interaction matrix is structured [e.g. due to guilds or trophic levels;
@loh_structure_2013]. Similarly, the separate generalized linear models
for each species can severely overfit in some cases [@lee_learning_2012]. The
full Markov network should thus be preferred to the approximations when
it is computationally tractable.

Compositional data only contains enough degrees of freedom to estimate one
interaction per species pair [@schmidt_modeling_2012], so none of these methods
can identify the exact nature of the pairwise interactions (e.g. which species
in a positively-associated pair is facilitating the other). To estimate
asymmetric interactions, such as commensalism or predation, ecologists could use
time series, behavioral observations, manipulative experiments, or natural
history. These other sources of information could also be used to augment the
likelihood function with an informative prior distribution, reducing ecologists'
error and uncertainty relative to Figure 3's results.

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they make many fewer errors than existing
approaches, and can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple answer to
the question of how competition should affect a species' overall prevalence,
which has important implications for community-level modeling
[@strong_ecological_1984]. Specifically, Equation 1 can be used to calculate the
expected prevalence of a species in the absence of biotic influences as
$e^\alpha/(e^{0} + e^\alpha)$. Competition's effect on prevalence can then be
estimated by comparing this value with the observed prevalence (e.g. comparing
Figure 2D with Figure 2C). This novel quantitative result conflicts with most of
our null models, which unreasonably assume that prevalence would be the exactly
same in the absence of competition as it is in the observed data
[@roughgarden_competition_1983].

Markov networks---particularly the Ising model for binary networks---are very
well understood, having been studied for nearly a century
[@cipra_introduction_1987]. Tapping into this framework would thus allow
ecologists to take advantage of into a vast set of existing discoveries and
techniques for dealing with indirect effects, stability, and alternative stable
states. Numerous extensions to the basic network are possible as well. For
example, the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would help incorporate
networks of biotic interactions into species distribution models
[@pollock_understanding_2014] and lead to a better understanding of the
interplay between biotic and abiotic effects on community structure.
Alternatively, models could allow one species to alter the relationship between
two other species [@tjelmeland_markov_1998; cf @bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
specific ecological processes like competition by rejecting a general-purpose
null [@gotelli_empirical_2009], especially when the test statistic is
effectively just a correlation coefficient (Figure 4B). When the non-null
backdrop is not controlled for, Type I error rates can skyrocket, the apparent
sign of the interaction can change, and null models can routinely produce
misleading inferences (Figure 1, Figure 4C, @gotelli_empirical_2009).

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3, scientists in this field could more than double
their explanatory power by switching from null models to Markov networks (or
increase it nearly as much with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for indirect
effects could clearly improve our understanding of species' direct effects on
one another and on community structure.

@@

\noindent \textbf{Figure 1: A.} A small network of three competing species.
The tree (top) tends not to co-occur with either of the two shrub species, as
indicated by the strongly negative coefficient linking them. The two shrub
species also compete with one another, but more weakly (circled coefficient).
**B.** In spite of the competitive interactions between the two shrub species,
their shared tendency to occur in locations without trees makes their occurrence
vectors positively correlated (circled). **C.** Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a Markov
network leads to correct identification of the negative shrub-shrub interaction
(circled).

\noindent \textbf{Figure 2: A.} A small Markov network, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta_{12}$ coefficient is consistent with competition between the two species.
**B.** The coefficients determine the probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and includes $\beta_{12}$ only when both are
present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
frequencies of all possible co-occurrence patterns between the two species of
interest, as calculated in the previous panel. **D.** Without competition (i.e.
with $\beta_{12}=0$, each species would occur more often.

\noindent \textbf{Figure 3:} Proportion of variance in interaction
coefficients explained by each method versus number of sampled locations across
the three simulation types.

\noindent \textbf{Figure 4: A.} The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates. **B.** The null
model's estimates typically matched the (negative) correlation coefficient,
after controlling for landscape size. **C.** For any given interaction strength,
the null model was much more likely to misclassify its sign with 95% confidence
than the Markov network was.
